---
title: "Analysis: Are Lower Tire Pressures Faster?"
author: "Mike Page"
date: "11/13/2024"
format: html
execute:
  echo: false
  warning: false
  error: false
---

```{r}
pkgload::load_all()
library(dplyr)
library(ggplot2)
library(gt)
library(ggridges)
options(pillar.sigfig = 4)
```

If lower tire pressures offer more grip and higher tire pressures roll faster, what pressure should we use? The true answer is it depends. On the track, the weather, the type of riding, the rider, the type of tire, and so on.

Rather than try to create a one size fits all answer, we can instead refine our question: are lower tire pressures faster than higher tire pressures on an enduro bike using downhill tires with inserts on a flow trail with tacky soil?

Answering this, and many questions like these can hep us build up an arsenal of knowledge we can use to tune our bikes to different scenarios.

### Testing setup

#### The bike

My test bike of choice was a Raaw Madonna v2.2. At 188cm in height, I ride a size large. The geometry of this bike feels great, with the high stack and well proportion chainstays creating intuitive handling which make placing weight between the tires feel effortless. A good thing when trying to assess their performance.

Michellin Wild Enduro Racing Line tires were chosen for this test. It is a tire I have a lot of experience on and believe it is well suited for the loose over hardpark type of terrain that is frequently found here in the south of Switzerland. Inside the tires is a set of cushcore pro inserts.

#### The track

The upper section of Mondini Downhill, a blue flow trail, was used for testing. It has a steep roll in followed by three flattish corners with just the right gradient that your tires begin to break traction if you don't pedal or touch the breaks.

Keeping the testing to only three corners allows us to reduce the scope other factors such as mistakes and fitness may have on the results. Simply put, the more times we can do a run, the more we can be sure the result is due to what we are testing, and not just to a bad run. The higher number of runs also allows us to perfect our lines and find the limits of traction.

Collectively, this setup allows us to create a consistent testing environment where we try to control the effects of pedalling, breaking, fitness and mistakes and isolate as much as we can the effects the tire pressures are having on performance.

```{=html}
<div class="TrailforksWidgetTrail" data-w="100%" data-h="350px" data-trailid="97411" data-activitytype="1" data-map="1" data-basemap="trailforks" data-elevation="1" data-photos="0" data-title="1" data-info="1" data-trail_opacity="25" data-v="2" data-basicmap="1"></div>

<script type="application/javascript" src="https://es.pinkbike.org/326/sprt/j/trailforks/iframeResizer.min.js"></script>
<script type="text/javascript">
var script = document.createElement("script"); script.setAttribute("src", "https://es.pinkbike.org/ttl-86400/sprt/j/trailforks/widget.js"); document.getElementsByTagName("head")[0].appendChild(script); var widgetCheck = false;
</script>
```

#### The experiment

Two sets of tire pressures were chosen for my 88kg weight: 28 (front/rear) PSI for the higher pressures and 18/20 (front/rear) for lower pressures. These pressures maintain a (nearly) consistent 0.9 front to rear ratio, with the high pressures aligning with recommendations such as [those from E13](https://support.ethirteen.com/hc/en-us/articles/360043258892-What-tire-pressure-should-I-run-with-your-2-4-2-5in-MTB-Tires).

After three warm up laps, each set of tire pressures was tested over 15 runs, resulting in 30 total laps. The runs were randomised into batches of 5 consecutive runs for a given tire pressure. This controls for the learning effect in which times will naturally get faster as you perfect the same track over and over (more on this later).

To time the runs, a [Freelap timing system](https://www.freelap.com/mtb/) was used, with a claimed accuracy of 0.02s. Conditions on the day were dry, with the soil being tacky after retaining some moisture from rain several days prior.

### The Results

If we look at the performance of the runs as time progresses, irrespective of what tire pressure was used, we can see that the times get progressively lower with each run. This is expected, given that a learning effect was presumably taking place as I learnt the track.

```{r}
plot <- results |>
  ggplot(aes(x = run, y = time_seconds)) +
  geom_point(size = 3, alpha = .8) +
  geom_smooth() +
  theme_minimal() +
  labs(
    title = "Performance improved as the number of runs increased",
    x = "Run number",
    y = "Time (s)"
  )

plotly_low_config(plot) |>
  config(
    displayModeBar = TRUE,
    displaylogo = FALSE,
    modeBarButtonsToRemove = list(
      "zoom",
      # "pan",
      "select",
      # "zoomIn",
      # "zoomOut",
      "autoScale",
      # "resetScale",
      "lasso2d",
      "hoverClosestCartesian",
      "hoverCompareCartesian"
    ),
    # Download button
    toImageButtonOptions = list(
      height = NULL,
      width = NULL,
      scale = 6
    )
  )
```

The fastest times of the day were 10.45 (s) on run 19 run 23, each with the higher air pressures. But if we look at the average times across all runs for each set of tire pressures, factor in that the timing system has an accuracy of 0.02s, we can see that the results are almost identical:

```{r}
results |>
  group_by(air_pressure_psi) |>
  summarise(average_time = mean(time_seconds)) |>
  arrange(average_time) |>
  rename(`Air Pressure (PSI)` = air_pressure_psi, `Average Time (s)` = average_time) |>
  gt() |>
  fmt_number()
```

But, averages can be deceiving. If we look at how consistent the times were for each of the sets of tire pressures, we can see the higher pressures showed much more vairance in performance, whereas the lower pressures were much more consistent:

```{r}
results |>
  ggplot(aes(x = time_seconds, y = air_pressure_psi, fill = air_pressure_psi)) +
  geom_density_ridges(scale = 4, alpha = .5) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_ridges() +
  scale_fill_viridis_d(option = "C", begin = .3, end = .8, guide = "none") +
  labs(
    title = "Lower pressures (purple) were more consistent. \nHigher pressures (yellow) showed greater variance.",
    y = "Air pressure front/rear (psi)",
    x = "Time (s)"
  )
```

As I was conducting the testing, I realised that as I switch to a new batch of five runs on a different set of tire pressures, it took me a couple of runs to adjust to the new settings. If we control for this by removing the results for the first two runs of each new batch, we can see that the results shift slightly in favour of the lower tire pressures:

```{r}
results |>
  filter(lap != 1 & lap != 2) |>
  group_by(air_pressure_psi) |>
  summarise(average_time = mean(time_seconds)) |>
  arrange(average_time)
```

Still small differences, but perhaps this time a meaningful difference if racing against the clock.

Examing the variance in the times with this adjust period controlled for, we see the same results as previously that higher tire pressures had more variance, with lower tire pressures being more consistent:

```{r}
results |>
  filter(lap != 1 & lap != 2) |>
  ggplot(aes(x = time_seconds, y = air_pressure_psi, fill = air_pressure_psi)) +
  geom_density_ridges(scale = 4, alpha = .5) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_ridges() +
  scale_fill_viridis_d(option = "C", begin = .3, end = .8, guide = "none") +
  labs(
    title = "Even after controlling for an adjustment period \nLower pressures were more consistent. \nHigher pressures showed greater variance.",
    y = "Air pressure front/rear (psi)",
    x = "Time (s)"
  )
```

### Conculstion

If you want consistent performance on tacky corners where your tires are breaking traction, lower tire pressures seem to be a smarter choice. But of course this speaks nothing to the downsides of slowing rolling resistance on climbs, how the same pressures might perform on other sections of track, and numerous other things.

My personal take away is perhaps I can afford to drop a little pressure from my tires. I've been riding 25/28 PSI with DH tires for a while now, and perhaps I can get more consistency and a little more dampening through dropping a few PSI.

What about you?